Working title: Regression models for Cylindrical data.

### 2018-07-02 Manuscript/ManuscriptMarkdown.tex
Manuscript Update

---
title: "Working title: Regression models for Cylindrical data."
author: |
  | Jolien Cremers^[Corresponding author: j.cremers@uu.nl] $^1$, Helena J.M. Pennings$^{2,3}$,  Christophe Ley $^{4}$
  | $^1$Department of Methodology and Statistics, Utrecht University
  | $^2$TNO
  | $^3$Department of Education, Utrecht University
  | $^4$Department of Applied Mathematics, Computer Science and Statistics, Ghent University

bibliography: CircularData.bib
csl: apa.csl
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: yes
  html_document: default
  word_document: default
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{multirow}
  - \usepackage{appendix}
  - \usepackage{color}
  - \usepackage{hyperref}
  - \usepackage{subcaption}
keywords: bla vla
documentclass: article
pandoc_args: --natbib
abstract: |
  This is the abstract
---
```{r inlude = FALSE, echo = FALSE}

knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r, echo = FALSE}
library(circular)
library(bpnreg)
library(haven)
library(tikzDevice)
library(plotrix)
library(tidyr)
library(dplyr) #sample_n()
library(MASS) #mvrnorm()
library(kableExtra) #latex tables
```

@@

Cylindrical data are data that consist of a linear component and a directional 
component (a set of angles). In this paper, the directional component is 
circular meaning that it consists of a single angle instead of a set of angles. 
Figure \ref{} \textcolor{red}{NOG MAKEN} shows the difference between a circular variable (left) and a 
linear variable (right). The most obviuous difference is that on a circular 
scale the datapoints 0$^\circ$ and 360$^\circ$ are connected and in fact 
represent the same number while on a linear scale the two ends, $-\infty$ and 
$\infty$ are not connected. Cylindrical data occur in several fields of 
research, such as for instance in meteorology [@garcia2013exploring], ecology 
[@garcia2014test] or marine research [@lagona2015hidden]. Several types of data 
in psychological research are also of a cylindrical nature. For example, in 
research on human navigation in the field of cognitive psychology both distance,
a linear variable, and direction, a circular variable, of movement are of 
interest [@chrastil2017rotational]. In eye-tracking research, which is used for 
investigating cognitive processes we can also speak of cylindrical data. When 
measuring eye-movements we speak of the eye-movements themselves, the saccades, 
and fixations, the periods of time between movements when the eyes are still and
looking at one point. Of the saccades both the direction, a circular variable, 
and the duration, a linear variable, are of interest (for a review of 
eye-tracking research see @rayner200935th). Data from circumplex measurement
instruments, e.g. the interpersonal circumplex as used in personality
psychology, are also of a cylindrical nature (see Section \ref{Example} for a
more detailed explanation). In this paper we will discuss how a correct
statistical treatment of such cylindrical data can lead to new insights. In
particular we will show how cylindrical models pave the way for circular-linear
and linear-circular regression. We will do this for a motivating example from
the field of educational psychology. We will refer to this example data as the
teacher data. In this example, apart from modelling the relation between the
linear and circular component of a cylindrical variable we would also like to
predict the two components from a set of covariates in a regression model. The
teacher data will be further introduced in Sections \ref{Example} and
\ref{DataAnalysis}.

As is the case for circular data, the analysis of cylindrical data requires 
special methods. For an overview of methods for circular data see
@mardia2000directional and @jammalamadaka2001topics and for modern treatments
@ley2017modern. In the literature several methods have been put forward to
model the relation between the linear and circular component of a cylindrical 
variable. Some of these are based on regressing the linear component onto the 
circular component using the following type of relation: $$y = \beta_0 + 
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,$$ where $y$ is the linear
component and $\theta$ the circular component [@mardia1978model;
@johnson1978some; @mastrantonio2015bayesian]. Others model the relation in a 
different way, e.g. by specifying a multivariate model for several linear and 
circular variables and modelling their covariance matrix 
[@mastrantonio2018joint] or by proposing a joint cylindrical distribution. 
For example, @abe2017tractable introduce a cylindrical distribution  based on a 
Weibull distribution for the linear component and a sine-skewed von Mises 
distribution for the circular component and link these through their shape and 
concentration parameters respectively. However, none of the methods that have 
been proposed thus far include additional covariates on which both the circular 
and linear component are regressed.

@@

The motivating example for this article comes from the field of educational 
psychology and was collected for the studies of @pennings2017interpersonal,
@Claessens2016side and @vanderWant2015role. These studies concern research on 
classroom climate. An indicator of classroom climate is the students'
perceptions of their teachers' interpersonal behavior. These interpersonal
perceptions, both in educational psychology as well as in other areas of
psychology, can be measured using circumplex measurement instruments (see
@horowitz2010handbook for an overview of many such instruments).

The Questionnaire on Teacher Interaction (QTI) [@wubbels2006interpersonal] is
one such circumplex measurement instrument that can be used to study student
perceptions of their teachers' interpersonal behavior. This instrument contains
items that load on two dimensions: Agency and Communion. Agency refers to the
degree of power or control a teacher exerts in interaction with his/her
students. Communion refers to the degree of friendliness or affiliation a
teacher conveys in interaction with his/her students. The loadings on the two 
dimensions of the QTI can be placed in a two-dimensional space formed by Agency
(vertical) and Communion (horizontal), see Figure \ref{QTI}. Different parts of
this space are characterized by different teacher behavior, e.g. 'helpful' or
'uncertain'. We call the two-dimensional space the interpersonal circle (IPC).
The idea is that the IPC is ``a continuous order with no beginning or end''
[@gurtman2009exploring, p. 2]. We call such ordering a circumplex ordering and
the IPC is therefore often called the interpersonal circumplex. The ordering
also implies that scores on the IPC could be viewed as a circular variable.

@@

@Cremers2018Assessing show how data from the IPC can be
considered circular data and analyzed as such using a regression model. The 
two-dimension scores Agency and Communion can be converted to a circular score 
using the two-argument arctangent function in Equation \eqref{PredVal}, where 
$A$ represents a score on the Agency dimension and $C$ represents a score on the
Communion dimension. The resulting circular variable $\theta$ can then be
modeled and for example be regressed onto covariates in a circular regression 
model. However, when two-dimensional data are converted to the circle we lose 
some information, namely the length of the two-dimensional vector $(A, C)$, 
\emph{i.e.}, its euclidean norm $\mid\mid (A, C) \mid\mid$. This length 
represents the strength of the type of interpersonal behavior a teacher shows 
towards his/her students. In a cylindrical model we are able to incorporate this
information, and model a circular variable $\theta$ together with a linear 
variable corresponding to $\mid\mid (A, C) \mid\mid$. This will lead to an 
improved analysis of data from the IPC.  In the next section we will introduce 
several models that we can use for a more accurate and more informative regression
analysis on the teacher data. More information on the teacher data will be given
in Section \ref{DataAnalysis}.

@@

```{r loaddata, return = FALSE, echo = FALSE}
Dat <- read_spss("Data Heleen/mergeddata.sav")

#only select the first measurement occasion
Dat <- subset(Dat, Time == 0)

#sort on class size
Dat <- Dat[order(-Dat$nklas), ]

#Now remove duplicates 
#(because dataframe is sorted on class size the smallest classes are removed)

duplicates <- duplicated(Dat[,c("docnr", "Time")])
Dat <- Dat[!duplicates,]

#center and remove missings for self-efficacy
Dat <- Dat %>% drop_na(Efficacy_SE)
Dat$SEc <- Dat$Efficacy_SE- mean(Dat$Efficacy_SE)

set.seed(50)

#Create holdout and test sample
Dat.hold <- sample_n(Dat, 15)
Dat.test <- Dat[-as.numeric(rownames(Dat.hold)),]

#Compute circular and linear outcome
theta.test <- atan2(Dat.test$lAgency, Dat.test$lcommunion)
y.test     <- sqrt(Dat.test$lAgency^2 + Dat.test$lcommunion^2)

theta.hold <- atan2(Dat.hold$lAgency, Dat.hold$lcommunion)
y.hold     <- sqrt(Dat.hold$lAgency^2 + Dat.hold$lcommunion^2)

```

@@

In this section we will introduce four cylindrical regression models. We will 
extend the models from @mastrantonio2018joint and @abe2017tractable to contain 
predictors for the linear and circular outcomes in the cylindrical model. 
Additionally we introduce two models where both variates $\Theta$ and $Y$ are
predicted by covariates and that model the relation between the linear and
circular outcome as follows (following @wang2015joint and
@mastrantonio2015bayesian):

@@

where $r$ will be introduced in Section \ref{CL-(G)PN}, the error term $\epsilon
\sim N(0, \sigma)$, $\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, 
\gamma_q$ are the intercept and regression coefficients and   $x_1, \dots, x_q$
are the $q$ covariate values. In this model $Y$ follows a normal distribution
and $\Theta$ a projected normal (PN) or general projected normal (GPN)
distribution on the circle.

@@

In both of these models the relation between the circular variable $\Theta \in
[0, 2\pi)$ and linear variable $Y \in (-\infty, + \infty)$ is specified as in
Equation \ref{circlinlink}. The distribution of $Y$ conditional on $\Theta$ is
then given as:

@@

where $c = \begin{bmatrix} r \cos \theta \\  r\sin \theta \end{bmatrix}^t 
\begin{bmatrix} \gamma_{cos} \\ \gamma_{sin} \end{bmatrix}$, $r \geq 0$, 
$\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, \gamma_q$ are the
intercept and regression coefficients and $\sigma^2 \geq 0$ is the error
variance. The linear outcome thus has a normal distribution, conditional on
$\Theta$.

For the circular outcome we assume either a projected normal (PN) or a
general projected normal (GPN) distribution. These distributions arise from the
radial projection of a distribution defined on the plane onto the circle. The
relation between a bivariate variate $\boldsymbol{S}$ in the plane and the
circular outcome $\Theta$ is defined as follows:

@@

where $R = \mid\mid S \mid\mid$, the euclidean norm of the bivariate vector 
$\boldsymbol{S}$. In the PN distribution we assume $\boldsymbol{S} \sim 
N_2(\boldsymbol{\mu}, \boldsymbol{I})$ and in the GPN we assume $\boldsymbol{S} 
\sim N_2(\boldsymbol{\mu}, \boldsymbol{\Sigma})$ where $\boldsymbol{\Sigma} = 
\begin{bmatrix} \tau^2 + \rho^2 & \rho\\ \rho & 1 \end{bmatrix}$, $\rho \in 
(-\infty, +\infty)$ and $\tau^2 \geq 0$ (as in @hernandez2016general).

Following @nunez2011bayesian, the joint density of $\Theta$ and $R$ for the PN
distribution in a regression set-up equals:

@@

In a regression setup the outcome $\theta_i$ for each individual $i = 1,
\dots, n$, where $n$ is the sample size, is generated independently from
Equation (\ref{pnreg}). The mean vector $\boldsymbol{\mu}_i \in \mathbb{R}^2$ is 
defined as $\boldsymbol{\mu}_i = \boldsymbol{z}_i\boldsymbol{B}$. The vector 
$\boldsymbol{z}_i$ is a a vector of $p$ covariate values and $\boldsymbol{B} = 
(\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})$ contain the regression 
coefficients. Note however that the dimensions of $\boldsymbol{\beta}^{I}$ and 
$\boldsymbol{\beta }^{I}$ need not necessarily be the same and we are thus 
allowed to have a different set of predictor variables and vectors 
$\boldsymbol{z}_i^I$ and $\boldsymbol{z}_i^{II}$ for the two components of 
$\boldsymbol{\mu}_i$.

Following @wang2012directional and @hernandez2016general the joint density of
$r$ and $\theta$ for the GPN distribution equals:

@@

where $\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \rho^2  & \rho\\ \rho & 1 
\end{bmatrix}$, $\boldsymbol{u}= \begin{bmatrix} \cos \theta \\ \sin \theta 
\end{bmatrix}$. In a regression setup the outcome $\theta_i$ for each individual
$i = 1, \dots, n$, where $n$ is the sample size, is generated independently
from Equation (\ref{gpnreg}). The mean vector $\boldsymbol{\mu}_i \in 
\mathbb{R}^2$ is defined as $\boldsymbol{\mu}_i = 
\boldsymbol{z}_i(\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})$. The vector 
$\boldsymbol{z}_i$ is a a vector of $p$ covariate values and each 
$\boldsymbol{\beta}^{k}$ is a vector with two regression coefficients, one for 
each of the two components of $\boldsymbol{\mu}_i$. Note that for the CL-GPN 
model we do need to have the same predictors for both components of 
$\boldsymbol{\mu}_i$.

Both cylindrical models introduced in this section are estimated using MCMC 
methods. These methods are based on @nunez2011bayesian, @wang2012directional and
@hernandez2016general for the regression of the circular outcome. A detailed
description of the Bayesian estimation and MCMC samplers can be found in
Appendices \ref{A1} and \ref{A2}.

@@

This model is an extension of the cylindrical model introduced by
@abe2017tractable to the regression context. The density of the model for the
pair of the circular variable $\theta \in [0, 2\pi)$ and linear variable $y$ on
the positive real half-line $[0, + \infty)$ is:

@@

In a regression setup the outcome vector $(\theta_i, y_i)^t$ for each individual
$i = 1, \dots, n$, where $n$ is the sample size, is generated independently from
Equation (\ref{WeiSSVMdensity}). The parameter $\alpha > 0$ is a linear shape 
parameter, $\beta_i = \exp(\boldsymbol{x}_i^t\boldsymbol{\nu}) > 0$ is a linear 
scale parameter, $\mu_i = \eta_0 + 
2\tan^{-1}(\boldsymbol{z}_i^t\boldsymbol{\eta}) \in [0, 2\pi)$ is a circular 
location parameter, $\kappa > 0$ is a cicular concentration parameter and 
$\lambda \in [-1, 1]$ is a circular skewness parameter. The parameter 
$\boldsymbol{\nu}$ is a vector of $q$ regression coefficients and intercept
$\nu_j \in (-\infty, +\infty)$ where $j = 0, \dots, q$, for the prediction of
$y$. The parameter $\eta_0 \in [0, 2\pi)$ is the intercept and
$\boldsymbol{\eta}$, is a vector of $p$ regression coefficients $\eta_j \in 
(-\infty, +\infty)$, where $j = 1, \dots, p$, for the prediction of $\theta$. 
The vector $\boldsymbol{x}_i$ is a vector of predictor values for the prediction
of $y$ and $\boldsymbol{z}_i$ is a vector of predictor values for the prediction
of $\theta$.

As in @abe2017tractable, the conditional distribution of $y$ given $\theta$ is a
Weibull distribution with scale $\alpha$ and shape
$\beta(1-\tanh(\kappa)\cos(\theta - \mu))^{1-\alpha}$ and the conditional 
distribution of $\theta$ given $y$ is a sine skewed von Mises distribution with 
location parameter $\mu$ and concentration parameter $(\beta 
y)^\alpha\tanh(\kappa)$.

@@

This model is an extension of the cylindrical model introduced by
@mastrantonio2018joint to the regression context. The model contains $p$
circular outcomes and $q$ linear outcomes. The circular outcomes 
$\boldsymbol{\Theta} = (\boldsymbol{\Theta}_1, \dots, \boldsymbol{\Theta}_p)$ 
are modeled together by a multivariate GPN distribution and the linear outcomes 
$\boldsymbol{Y} = (\boldsymbol{Y}_1, \dots, \boldsymbol{Y}_q)$  are modeled 
together by a multivariate skew normal distribution [@sahu2003new]. Because the
GPN distribution is modelled using a so-called augmented representation (as in
(\ref{projection}) and (\ref{gpnreg})) it is convenient to use a similar tactic
for modelling the multivariate skew normal distribution. As in
@mastrantonio2018joint we use the following representation of the linear
outcomes:

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Working title: Regression models for Cylindrical data.},
            pdfkeywords={bla vla},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Working title: Regression models for Cylindrical data.}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Jolien Cremers\footnote{Corresponding author:
  \href{mailto:j.cremers@uu.nl}{\nolinkurl{j.cremers@uu.nl}}} \(^1\),
Helena J.M. Pennings\(^{2,3}\), Christophe Ley \(^{4}\)\\
\(^1\)Department of Methodology and Statistics, Utrecht University\\
\(^2\)TNO\\
\(^3\)Department of Education, Utrecht University\\
\(^4\)Department of Applied Mathematics, Computer Science and
Statistics, Ghent University}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \date{}
  \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

\usepackage{multirow}
\usepackage{appendix}
\usepackage{color}
\usepackage{hyperref}
\usepackage{subcaption}

\begin{document}
\maketitle
\begin{abstract}
This is the abstract
\end{abstract}

@@

Cylindrical data are data that consist of a linear component and a
directional component (a set of angles). In this paper, the directional
component is circular meaning that it consists of a single angle instead
of a set of angles. Figure \ref{} \textcolor{red}{NOG MAKEN} shows the
difference between a circular variable (left) and a linear variable
(right). The most obviuous difference is that on a circular scale the
datapoints 0\(^\circ\) and 360\(^\circ\) are connected and in fact
represent the same number while on a linear scale the two ends,
\(-\infty\) and \(\infty\) are not connected. Cylindrical data occur in
several fields of research, such as for instance in meteorology
(García-Portugués, Crujeiras, \& González-Manteiga, 2013), ecology
(García-Portugués, Barros, Crujeiras, González-Manteiga, \& Pereira,
2014) or marine research (Lagona, Picone, Maruotti, \& Cosoli, 2015).
Several types of data in psychological research are also of a
cylindrical nature. For example, in research on human navigation in the
field of cognitive psychology both distance, a linear variable, and
direction, a circular variable, of movement are of interest (Chrastil \&
Warren, 2017). In eye-tracking research, which is used for investigating
cognitive processes we can also speak of cylindrical data. When
measuring eye-movements we speak of the eye-movements themselves, the
saccades, and fixations, the periods of time between movements when the
eyes are still and looking at one point. Of the saccades both the
direction, a circular variable, and the duration, a linear variable, are
of interest (for a review of eye-tracking research see Rayner (2009)).
Data from circumplex measurement instruments, e.g.~the interpersonal
circumplex as used in personality psychology, are also of a cylindrical
nature (see Section \ref{Example} for a more detailed explanation). In
this paper we will discuss how a correct statistical treatment of such
cylindrical data can lead to new insights. In particular we will show
how cylindrical models pave the way for circular-linear and
linear-circular regression. We will do this for a motivating example
from the field of educational psychology. We will refer to this example
data as the teacher data. In this example, apart from modelling the
relation between the linear and circular component of a cylindrical
variable we would also like to predict the two components from a set of
covariates in a regression model. The teacher data will be further
introduced in Sections \ref{Example} and \ref{DataAnalysis}.

As is the case for circular data, the analysis of cylindrical data
requires special methods. For an overview of methods for circular data
see Mardia \& Jupp (2000) and Jammalamadaka \& Sengupta (2001) and for
modern treatments Ley \& Verdebout (2017). In the literature several
methods have been put forward to model the relation between the linear
and circular component of a cylindrical variable. Some of these are
based on regressing the linear component onto the circular component
using the following type of relation: \[y = \beta_0 + 
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is
the linear component and \(\theta\) the circular component (Johnson \&
Wehrly, 1978; Mardia \& Sutton, 1978; Mastrantonio, Maruotti, \&
Jona-Lasinio, 2015). Others model the relation in a different way,
e.g.~by specifying a multivariate model for several linear and circular
variables and modelling their covariance matrix (Mastrantonio, 2018) or
by proposing a joint cylindrical distribution. For example, Abe \& Ley
(2017) introduce a cylindrical distribution based on a Weibull
distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their
shape and concentration parameters respectively. However, none of the
methods that have been proposed thus far include additional covariates
on which both the circular and linear component are regressed.

@@

The motivating example for this article comes from the field of
educational psychology and was collected for the studies of Pennings et
al. (2017), Claessens (2016) and Want (2015). These studies concern
research on classroom climate. An indicator of classroom climate is the
students' perceptions of their teachers' interpersonal behavior. These
interpersonal perceptions, both in educational psychology as well as in
other areas of psychology, can be measured using circumplex measurement
instruments (see Horowitz \& Strack (2011) for an overview of many such
instruments).

The Questionnaire on Teacher Interaction (QTI) (Wubbels, Brekelmans,
Brok, \& Tartwijk, 2006) is one such circumplex measurement instrument
that can be used to study student perceptions of their teachers'
interpersonal behavior. This instrument contains items that load on two
dimensions: Agency and Communion. Agency refers to the degree of power
or control a teacher exerts in interaction with his/her students.
Communion refers to the degree of friendliness or affiliation a teacher
conveys in interaction with his/her students. The loadings on the two
dimensions of the QTI can be placed in a two-dimensional space formed by
Agency (vertical) and Communion (horizontal), see Figure \ref{QTI}.
Different parts of this space are characterized by different teacher
behavior, e.g. `helpful' or `uncertain'. We call the two-dimensional
space the interpersonal circle (IPC). The idea is that the IPC is ``a
continuous order with no beginning or end'' (Gurtman, 2009, p. 2). We
call such ordering a circumplex ordering and the IPC is therefore often
called the interpersonal circumplex. The ordering also implies that
scores on the IPC could be viewed as a circular variable.

@@

Cremers et al. (2018a) show how data from the IPC can be considered
circular data and analyzed as such using a regression model. The
two-dimension scores Agency and Communion can be converted to a circular
score using the two-argument arctangent function in Equation
\eqref{PredVal}, where \(A\) represents a score on the Agency dimension
and \(C\) represents a score on the Communion dimension. The resulting
circular variable \(\theta\) can then be modeled and for example be
regressed onto covariates in a circular regression model. However, when
two-dimensional data are converted to the circle we lose some
information, namely the length of the two-dimensional vector \((A, C)\),
\emph{i.e.}, its euclidean norm \(\mid\mid (A, C) \mid\mid\). This
length represents the strength of the type of interpersonal behavior a
teacher shows towards his/her students. In a cylindrical model we are
able to incorporate this information, and model a circular variable
\(\theta\) together with a linear variable corresponding to
\(\mid\mid (A, C) \mid\mid\). This will lead to an improved analysis of
data from the IPC. In the next section we will introduce several models
that we can use for a more accurate and more informative regression
analysis on the teacher data. More information on the teacher data will
be given in Section \ref{DataAnalysis}.

@@

@@

In this section we will introduce four cylindrical regression models. We
will extend the models from Mastrantonio (2018) and Abe \& Ley (2017) to
contain predictors for the linear and circular outcomes in the
cylindrical model. Additionally we introduce two models where both
variates \(\Theta\) and \(Y\) are predicted by covariates and that model
the relation between the linear and circular outcome as follows
(following Wang, Gelfand, \& Jona-Lasinio (2015) and Mastrantonio et al.
(2015)):

@@

where \(r\) will be introduced in Section \ref{CL-(G)PN}, the error term
\(\epsilon \sim N(0, \sigma)\),
\(\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, \gamma_q\) are
the intercept and regression coefficients and \(x_1, \dots, x_q\) are
the \(q\) covariate values. In this model \(Y\) follows a normal
distribution and \(\Theta\) a projected normal (PN) or general projected
normal (GPN) distribution on the circle.

@@

In both of these models the relation between the circular variable
\(\Theta \in [0, 2\pi)\) and linear variable
\(Y \in (-\infty, + \infty)\) is specified as in Equation
\ref{circlinlink}. The distribution of \(Y\) conditional on \(\Theta\)
is then given as:

@@

where
\(c = \begin{bmatrix} r \cos \theta \\ r\sin \theta \end{bmatrix}^t \begin{bmatrix} \gamma_{cos} \\ \gamma_{sin} \end{bmatrix}\),
\(r \geq 0\),
\(\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, \gamma_q\) are
the intercept and regression coefficients and \(\sigma^2 \geq 0\) is the
error variance. The linear outcome thus has a normal distribution,
conditional on \(\Theta\).

For the circular outcome we assume either a projected normal (PN) or a
general projected normal (GPN) distribution. These distributions arise
from the radial projection of a distribution defined on the plane onto
the circle. The relation between a bivariate variate \(\boldsymbol{S}\)
in the plane and the circular outcome \(\Theta\) is defined as follows:

@@

where \(R = \mid\mid S \mid\mid\), the euclidean norm of the bivariate
vector \(\boldsymbol{S}\). In the PN distribution we assume
\(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{I})\) and in the
GPN we assume
\(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{\Sigma})\) where
\(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \rho^2 & \rho\\ \rho & 1 \end{bmatrix}\),
\(\rho \in (-\infty, +\infty)\) and \(\tau^2 \geq 0\) (as in
Hernandez-Stumpfhauser, Breidt, \& Woerd (2016)).

Following Nuñez-Antonio, Gutiérrez-Peña, \& Escarela (2011), the joint
density of \(\Theta\) and \(R\) for the PN distribution in a regression
set-up equals:

@@

In a regression setup the outcome \(\theta_i\) for each individual
\(i = 1, \dots, n\), where \(n\) is the sample size, is generated
independently from Equation (\ref{pnreg}). The mean vector
\(\boldsymbol{\mu}_i \in \mathbb{R}^2\) is defined as
\(\boldsymbol{\mu}_i = \boldsymbol{z}_i\boldsymbol{B}\). The vector
\(\boldsymbol{z}_i\) is a a vector of \(p\) covariate values and
\(\boldsymbol{B} = (\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})\)
contain the regression coefficients. Note however that the dimensions of
\(\boldsymbol{\beta}^{I}\) and \(\boldsymbol{\beta }^{I}\) need not
necessarily be the same and we are thus allowed to have a different set
of predictor variables and vectors \(\boldsymbol{z}_i^I\) and
\(\boldsymbol{z}_i^{II}\) for the two components of
\(\boldsymbol{\mu}_i\).

Following Wang \& Gelfand (2013) and Hernandez-Stumpfhauser et al.
(2016) the joint density of \(r\) and \(\theta\) for the GPN
distribution equals:

@@

where
\(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \rho^2 & \rho\\ \rho & 1 \end{bmatrix}\),
\(\boldsymbol{u}= \begin{bmatrix} \cos \theta \\ \sin \theta \end{bmatrix}\).
In a regression setup the outcome \(\theta_i\) for each individual
\(i = 1, \dots, n\), where \(n\) is the sample size, is generated
independently from Equation (\ref{gpnreg}). The mean vector
\(\boldsymbol{\mu}_i \in \mathbb{R}^2\) is defined as
\(\boldsymbol{\mu}_i = \boldsymbol{z}_i(\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})\).
The vector \(\boldsymbol{z}_i\) is a a vector of \(p\) covariate values
and each \(\boldsymbol{\beta}^{k}\) is a vector with two regression
coefficients, one for each of the two components of
\(\boldsymbol{\mu}_i\). Note that for the CL-GPN model we do need to
have the same predictors for both components of \(\boldsymbol{\mu}_i\).

Both cylindrical models introduced in this section are estimated using
MCMC methods. These methods are based on Nuñez-Antonio et al. (2011),
Wang \& Gelfand (2013) and Hernandez-Stumpfhauser et al. (2016) for the
regression of the circular outcome. A detailed description of the
Bayesian estimation and MCMC samplers can be found in Appendices
\ref{A1} and \ref{A2}.

@@

This model is an extension of the cylindrical model introduced by Abe \&
Ley (2017) to the regression context. The density of the model for the
pair of the circular variable \(\theta \in [0, 2\pi)\) and linear
variable \(y\) on the positive real half-line \([0, + \infty)\) is:

@@

In a regression setup the outcome vector \((\theta_i, y_i)^t\) for each
individual \(i = 1, \dots, n\), where \(n\) is the sample size, is
generated independently from Equation (\ref{WeiSSVMdensity}). The
parameter \(\alpha > 0\) is a linear shape parameter,
\(\beta_i = \exp(\boldsymbol{x}_i^t\boldsymbol{\nu}) > 0\) is a linear
scale parameter,
\(\mu_i = \eta_0 + 2\tan^{-1}(\boldsymbol{z}_i^t\boldsymbol{\eta}) \in [0, 2\pi)\)
is a circular location parameter, \(\kappa > 0\) is a cicular
concentration parameter and \(\lambda \in [-1, 1]\) is a circular
skewness parameter. The parameter \(\boldsymbol{\nu}\) is a vector of
\(q\) regression coefficients and intercept
\(\nu_j \in (-\infty, +\infty)\) where \(j = 0, \dots, q\), for the
prediction of \(y\). The parameter \(\eta_0 \in [0, 2\pi)\) is the
intercept and \(\boldsymbol{\eta}\), is a vector of \(p\) regression
coefficients \(\eta_j \in (-\infty, +\infty)\), where
\(j = 1, \dots, p\), for the prediction of \(\theta\). The vector
\(\boldsymbol{x}_i\) is a vector of predictor values for the prediction
of \(y\) and \(\boldsymbol{z}_i\) is a vector of predictor values for
the prediction of \(\theta\).

As in Abe \& Ley (2017), the conditional distribution of \(y\) given
\(\theta\) is a Weibull distribution with scale \(\alpha\) and shape
\(\beta(1-\tanh(\kappa)\cos(\theta - \mu))^{1-\alpha}\) and the
conditional distribution of \(\theta\) given \(y\) is a sine skewed von
Mises distribution with location parameter \(\mu\) and concentration
parameter \((\beta y)^\alpha\tanh(\kappa)\).

@@

This model is an extension of the cylindrical model introduced by
Mastrantonio (2018) to the regression context. The model contains \(p\)
circular outcomes and \(q\) linear outcomes. The circular outcomes
\(\boldsymbol{\Theta} = (\boldsymbol{\Theta}_1, \dots, \boldsymbol{\Theta}_p)\)
are modeled together by a multivariate GPN distribution and the linear
outcomes
\(\boldsymbol{Y} = (\boldsymbol{Y}_1, \dots, \boldsymbol{Y}_q)\) are
modeled together by a multivariate skew normal distribution (Sahu, Dey,
\& Branco, 2003). Because the GPN distribution is modelled using a
so-called augmented representation (as in (\ref{projection}) and
(\ref{gpnreg})) it is convenient to use a similar tactic for modelling
the multivariate skew normal distribution. As in Mastrantonio (2018) we
use the following representation of the linear outcomes:
